---
layout: post
title: "Visual meditations on house prices, Part 7: streams and recursion"
author: "Len Kiefer"
date: "2016-12-03"
summary: "Charts and graphs exploring house price trends"
group: navigation
theme :
  name : lentheme
---
# Introduction

I HAVE BEEN WATCHING SOME VIDEOS of [Plotcon 2016](https://plotcon.plot.ly/).  Most of the videos are worth watching ([check out the playlist](https://www.youtube.com/playlist?list=PLR7d32uh__xbrzSdfzxxCxCEFQnt7BcOj)), but I was particularly interested in this one from Hadley Wickham:

<iframe width="560" height="315" src="https://www.youtube.com/embed/cU0-NrUxRw4?list=PLR7d32uh__xbrzSdfzxxCxCEFQnt7BcOj" frameborder="0" allowfullscreen></iframe>

Among other things Hadley talks about the idea of nesting models and model results within a data frame.  That idea struck me as something that could be quite useful. Let's explore using house prices.

I've been collecting various graphical thoughts about house prices in my Visual Meditations series. For reference, the prior meditations are listed below, and I'll keep an updated list of all of them [here]({% post_url 2016-05-08-visual-meditations-on-house-prices %}).


* [Part 1: data wrangling ]({% post_url 2016-05-08-visual-meditations-on-house-prices-part1 %})
* [Part 2: sparklines and dots (animated) ]({% post_url 2016-05-08-visual-meditations-on-house-prices-part2 %})
* [Part 3: bubbles and bounce ]({% post_url 2016-05-10-visual-meditations-on-house-prices-part3 %})
* [Part 4: graph gallery ]({% post_url 2016-05-14-visual-meditations-on-house-prices-part4 %})
* [Part 5: distributions ]({% post_url 2016-08-13-visual-meditations-on-house-prices-part5 %})
* [Part 6: state recovery ]({% post_url 2016-11-03-visual-meditations-on-house-prices-part6 %})

These visualizations will be made in R, and I'll post code for some of the graphs at the bottom.


# Don't cross the streams

Before we get to recursion, let's think about streams.  On Friday I came up with some accidental art:

<blockquote class="twitter-tweet" data-lang="en"><p lang="en" dir="ltr">made an unintentional sorta streamgraph <a href="https://twitter.com/hashtag/graphfail?src=hash">#graphfail</a> <a href="https://t.co/uW4IOnbrcL">pic.twitter.com/uW4IOnbrcL</a></p>&mdash; Leonard Kiefer (@lenkiefer) <a href="https://twitter.com/lenkiefer/status/804448317891457024">December 1, 2016</a></blockquote>
<script async src="//platform.twitter.com/widgets.js" charset="utf-8"></script>

First we'll recreate this useless graph and see if we can get something out of it.

## The data

Once [again]({% post_url 2016-05-08-visual-meditations-on-house-prices-part1 %}) we'll use the [Freddie Mac House Price Index](http://www.freddiemac.com/finance/house_price_index.html) for many of these visualizations.  We're going to need a text file organized as described in that post. Just follow those examples to set up the data. Or you can download the two text files below:

1. [*state and national called fmhpi2016q3.txt*]({{ site.url }}/img/charts_nov_3_2016/fmhpi2016q3.txt)
2. [*metro called fmhpi2016q3metro.txt*]({{ site.url }}/img/charts_nov_3_2016/fmhpi2016q3metro.txt)

The code below will load the data and do some manipulations to generate the required variables.

``` {r, cache=FALSE,echo=T,message=FALSE,warnings=FALSE }
#Load some packages
library(data.table, warn.conflicts = FALSE, quietly=TRUE)
library(ggplot2, warn.conflicts = FALSE, quietly=TRUE)
library(dplyr, warn.conflicts = FALSE, quietly=TRUE)
library(zoo, warn.conflicts = FALSE, quietly=TRUE)
library(ggrepel, warn.conflicts = FALSE, quietly=TRUE)
library(ggthemes, warn.conflicts = FALSE, quietly=TRUE)
library(scales, warn.conflicts = FALSE, quietly=TRUE)
library(animation, warn.conflicts = FALSE, quietly=TRUE)
library(grid, warn.conflicts = FALSE, quietly=TRUE)
library(tidyr, warn.conflicts = FALSE, quietly=TRUE)
library(rgeos)
library(maptools)
library(albersusa)

d.metro <- fread("data/fmhpi2016q3metro.txt")
d.metro$date<-as.Date(d.metro$date, format="%m/%d/%Y")

#Now uses some data table caclulations to compute percent changes in house prices by state/metro
d.metro[,hpa12:=c(rep(NA,12),((1+diff(hpi,12)/hpi))^1)-1,by=metro]  
# set up statecode for primary state, first 2 digits after column in metro name
d.metro[,statecode:=substr(regmatches(metro,regexec(", +[A-Z][A-Z]",metro)),3,4)]

#load file with regions
region<-fread("data/region.txt")
reg.list<-unique(region[order(region),]$region) #list of regions
d.metro<-merge(d.metro,region,by="statecode")

# create quantiles across regions and dates:
d.metro[, hpa12.rmin:=quantile(hpa12,0,na.rm=T),by=c("region","date")]
d.metro[, hpa12.r5:=quantile(hpa12,.05,na.rm=T),by=c("region","date")]
d.metro[, hpa12.r25:=quantile(hpa12,.25,na.rm=T),by=c("region","date")]
d.metro[, hpa12.r50:=quantile(hpa12,.5,na.rm=T),by=c("region","date")]
d.metro[, hpa12.r75:=quantile(hpa12,.75,na.rm=T),by=c("region","date")]
d.metro[, hpa12.r95:=quantile(hpa12,.95,na.rm=T),by=c("region","date")]
d.metro[, hpa12.rmax:=quantile(hpa12,1,na.rm=T),by=c("region","date")]
```

Now we're ready to create the plot above.  Even with the labels you don't really get a lot out of it, but it's kind of pretty.


```{r ,fig-hpimed6-viz1, warning=FALSE, cache=TRUE,echo=T, message=FALSE,,fig.height=5,fig.width=7}
ggplot(data=d.metro,aes(x=date,fill=region,color=region))+
  geom_ribbon(aes(ymin=hpa12.rmin,ymax=hpa12.r5),alpha=0.1)+
  geom_ribbon(aes(ymin=hpa12.r5,ymax=hpa12.r25),alpha=0.5)+
  geom_ribbon(aes(ymin=hpa12.r25,ymax=hpa12.r75),alpha=0.75)+
  geom_ribbon(aes(ymin=hpa12.r75,ymax=hpa12.r95),alpha=0.5)+
  geom_ribbon(aes(ymin=hpa12.r95,ymax=hpa12.rmax),alpha=0.1)+
  geom_line(aes(y=hpa12.r50))+ theme_minimal()+
  theme(legend.position="none",plot.caption=element_text(hjust=0),
        plot.subtitle=element_text(face="italic"))+
  scale_y_continuous(label=percent)+
      coord_cartesian(xlim=c(as.Date("1990-01-01"),as.Date("2016-12-31")))+
  labs(x="",y="Annual House Price Percent Change",title="Distribution of year-over-year house price growth across metros", subtitle="Black line median metro, central region 25th to 75th percentiles,\nlighter regions are 5th to 25th (75th to 95th) and min to 5th (95th to max)",
       caption="@lenkiefer Source: Freddie Mac House Price Index (NSA), metros assigned to region based on primary state.")
```

What I wanted to do with this graph was compare the dispersion across metro areas of house price growth.  By adding a simple facet_wrap() statement we can get something a little more useful:

```{r ,fig-hpimed6-viz2, warning=FALSE, cache=TRUE,echo=T, message=FALSE,,fig.height=5,fig.width=7}
ggplot(data=d.metro,aes(x=date,fill=region,color=region))+
  geom_ribbon(aes(ymin=hpa12.rmin,ymax=hpa12.r5),alpha=0.1)+
  geom_ribbon(aes(ymin=hpa12.r5,ymax=hpa12.r25),alpha=0.5)+
  geom_ribbon(aes(ymin=hpa12.r25,ymax=hpa12.r75),alpha=0.75)+
  geom_ribbon(aes(ymin=hpa12.r75,ymax=hpa12.r95),alpha=0.5)+
  geom_ribbon(aes(ymin=hpa12.r95,ymax=hpa12.rmax),alpha=0.1)+
  geom_line(aes(y=hpa12.r50))+ theme_minimal()+
  facet_wrap(~region)+
  theme(legend.position="none",plot.caption=element_text(hjust=0),
        plot.subtitle=element_text(face="italic"))+
  scale_y_continuous(label=percent)+
      coord_cartesian(xlim=c(as.Date("1990-01-01"),as.Date("2016-12-31")))+
  labs(x="",y="Annual House Price Percent Change",title="Distribution of year-over-year house price growth across metros", subtitle="Black line median metro, central region 25th to 75th percentiles,\nlighter regions are 5th to 25th (75th to 95th) and min to 5th (95th to max)",
       caption="@lenkiefer Source: Freddie Mac House Price Index (NSA), metros assigned to region based on primary state.")
```
