---
  layout: post
title: "Economic and Housing Recap: January 2017"
author: "Len Kiefer"
date: "2017-01-30"
summary: "R statistics dataviz plotly housing mortgage data"
group: navigation
theme :
  name : lentheme
---

WE ARE ONE MONTH INTO 2017 AND IT IS TIME to recap the key economic and housing market releases for this month.  Much of the data we got this month was actually for December or the fourth quarter of 2016.  These data by and large matched our expectations as I [outlined in my 2016 year-in-review]({{ site.url}}/chartbooks/dec2016/index.html).

For this post I want to integrate my commentary with some graphs (made with [R](https://www.r-project.org/)) and the code I used to create these graphs. To make this analysis easy, we'll use the  [quantmod](https://cran.r-project.org/web/packages/quantmod/index.html) package to get the data from the St. Louis Federal Reserve Economic Database ([FRED](https://fred.stlouisfed.org/)).

# Load some data

The code below organizes our data.  I hacked this together using the quantmod package and the `getSymbols` function. This isn't the most elegant solution, but I don't use quantmod very often so this will have to do for now.  Basically I grab a series from FRED, captured in the *tickers* vector. Then I use a couple functions to generate a data table I can use in my graphs.

I'm sure there is a better way to do this, but this seems to work.

```{r, eval=T, message=F,warning=F}
library(tidyverse)
library(quantmod)

tickers<-c('EXHOSLUSM495S', # existing home sales
           'HSN1F',         # new home sales
           'HOUST1F',       # housing starts (total 1-unit)
           'HOUST',         # housing starts (total private)
           'HPIPONM226S',   # FHFA Purchase only house price index
           'PCE',           # Personal Consumption Expenditures
           'PI')            # Personal Income


make.data<-function(x){
df = getSymbols(x,src='FRED', auto.assign=F) 
dt = data.table(date=time(df), coredata(df) )
return(dt)
}

make.df<-function(i,indata=y){
  y<-make.data(tickers[i])
  out<-merge(indata,y,by="date",all=T)
  return(out)
  }
#initialize
my.dt<-make.data(tickers[1])

for (i in 2:length(tickers)){
  my.dt<-make.df(i,my.dt)
  }
```

  
RECENTLY I [HAVE BEEN]({% post_url 2017-01-22-build-flex %}) EXPLORING [FLEXDASHBOARDS](http://rmarkdown.rstudio.com/flexdashboard/index.html) to visualize data.  In this post I want to focus on a tool I've found particularly useful, [plotly]( https://plot.ly/r/).

Plotly enables you to make interactive [html widgets]( http://www.htmlwidgets.org/) that you can embed in your webpage or view from within [R](https://www.r-project.org/). I've been having a lot of fun converting existing visualizations I have made with [ggplot2](http://ggplot2.org/) into plotly visualizations using [ggplotly](https://plot.ly/ggplot2/).

In this post, let me share some of what I've been doing.
  
## The Plan

I'm going to include the code and discussion for several graphs I've been using.  I will use updated data that we used in our [Cross talk dashboard]({% post_url 2017-01-16-cross-talk-dashboard %}). These data cover weekly mortgage rates and house prices.

### The data

We've used these data before, but if you want to follow along they are:

* Mortgage rate data [rates.xlsx spreadsheet]({{ site.url}}/chartbooks/jan2017/rates.xlsx)
* Metro hpi files [hpimetro.csv]({{ site.url}}/chartbooks/jan2017/data/hpimetro.csv)
* State house price file [hpistate.csv]({{ site.url}}/chartbooks/jan2017/data/hpistate.csv)
* National house price file [hpiusa.csv]({{ site.url}}/chartbooks/jan2017/data/hpiusa.csv)

Now that if you store the data in a folder called *data*, you can prep the data with the following code:

```{r setup-data, cache=F,include=T}
library(tidyverse,quietly=T)
library(data.table,quietly=T)
library(htmlTable,quietly=T)
library(viridis,quietly=T)
library(DT,quietly=T)
library(plotly,quietly=T)
library(scales,quietly=T)
library(maps,quietly=T)
library(crosstalk,quietly=T)


####################
####  Load Data ####
####################

states_map <- map_data("state") # state data for map

# Load metro data
df<-fread("data/hpimetro.csv")
df$date<-as.Date(df$date, format="%m/%d/%Y")

# Load state data
df.state<-fread("data/hpistate.csv")
df.state$date<-as.Date(df.state$date, format="%m/%d/%Y")


# Load US data
df.us<-fread("data/hpiusa.csv")
df.us$date<-as.Date(df.us$date, format="%m/%d/%Y")

# Set up metro data for cross talk:
df.metro<-group_by(df,geo)
sd.metro <- SharedData$new(df.metro, ~geo)


#### Load Mortgage Rates Data
#### See for discussion http://lenkiefer.com/2017/01/08/mortgage-rate-viewer

####################################################################################################

dt<- read_excel('data/rates.xlsx',sheet= 'rates')
dt$date<-as.Date(dt$date, format="%m/%d/%Y")
dt<-data.table(dt) 
dt$year<-year(dt$date) # create year variable
dt[,week:=1:.N,by=year] # add a week number variable for week of year

```
  
# Mortgage rate plots

I've made many mortgage rates plots (see [here for 10 visualizations]({% post_url 2016-12-08-10-ways-to-visualize-rates %}) and 
[here for a flexdashboard]({% post_url 2017-01-08-mortgage-rate-viewer %})).

## Simple line plot

Let's start with a simple line plot:

```{r jan-29-2017-rate-graph-1}
g1<-
  ggplot(data=dt[year>2010,],aes(x=date,y=rate30,label=rate30))+geom_line()+
  geom_text(data=tail(dt,1),nudge_x=60,nudge_y=.1,color="red",size=2.5,hjust=0)+
  geom_point(data=tail(dt,1),size=2,color="red",alpha=0.75)+
  theme_minimal()+
  geom_hline(linetype=2,color="red",yintercept=tail(dt,1)$rate30)+
  labs(x="", y="",
       title="30-year Fixed Mortgage Rate (%)",
       subtitle="weekly rates since 2011",
       caption="@lenkiefer Source: Freddie Mac Primary Mortgage Market Survey")+
  theme(plot.title=element_text(size=18),
        plot.caption=element_text(hjust=0,vjust=1,margin=margin(t=10)),
        plot.background=element_rect(fill="#fffff8",color=NA))
g1

```

Converting this simple plot to plotly is quite easy.  We simply have to use `ggplotly()` with the graph name (g1) as the function argument.  If we want to add a range slider to select dates, we can add it with pipes via `%>% rangeslider()`.

```{r jan-29-2017-rate-graph-1-plotly, include=T}
ggplotly(g1) %>% rangeslider()
```

The graph above is a static screenshot, but you can see the interactive version by [clicking here]({{ site.url}}/chartbooks/jan2017/plotly/rates-jan29-2017-1.html).

## Distribution dot plot

I like to look at the distribution of observations.

```{r jan-29-2017-rate-graph-2}
g2<-
 ggplot(data=dt[year>2000,],aes(x=year,y=rate30,color=rate30,group=year))+
  geom_jitter(height=0,width=0.2,alpha=0.1,color="gray")+
  geom_jitter(height=0,width=0.2,alpha=0.82)+
  scale_x_continuous(breaks=seq(2001,2017,1))+
  theme_minimal()+
  scale_color_viridis(direction=-1,name="rate in pp")+
  labs(y="30-year fixed mortgage rate (%)",        x="year",
       title="Distribution of weekly mortgage rates",
       subtitle="Each dot average rate for a week",
       caption="@lenkiefer Source: Freddie Mac Primary Mortgage Market Survey")+
  theme(legend.position="none",plot.subtitle=element_text(face="italic"),
        plot.title=element_text(face="bold"),
        plot.caption=element_text(hjust=0))
g2
```

In this plot we group each year and then jitter the x position of each dot. By comparing the mass of points, we can get a feeling about how the distribution of weekly rates has shifted by year. 

### Add animation

With plotly, we can easily add animation to this plot to see how the distribution has evolved over time. We have to modify our original code slightly:

```{r jan-29-2017-rate-graph-2-plotly}
g2<-
 ggplot(data=dt[year>2000,],aes(x=year,y=rate30,color=rate30,group=year))+
  geom_jitter(height=0,width=0.2,alpha=0.1,color="gray")+
  geom_jitter(height=0,width=0.2,alpha=0.82,aes(frame=year))+
  scale_x_continuous(breaks=seq(2001,2017,1))+
  theme_minimal()+
  scale_color_viridis(direction=-1,name="rate in pp")+
  labs(y="30-year fixed mortgage rate (%)",        x="year",
       title="Distribution of weekly mortgage rates<br><i>Each dot a week</i><br>@lenkiefer Source: Primary Mortgage Market Survey",
       subtitle="Each dot average rate for a week",
       caption="@lenkiefer Source: Freddie Mac Primary Mortgage Market Survey")+
  theme(legend.position="none",plot.subtitle=element_text(face="italic"),
        plot.title=element_text(face="bold"),
        plot.caption=element_text(hjust=0))
ggplotly(g2) %>% animation_opts(frame=1000,transition=600,redraw=T)  

```

Once again, the graph above is a static screenshot, but you can see the interactive version by [clicking here]({{ site.url}}/chartbooks/jan2017/plotly/rates-jan29-2017-2.html).

# House Price visualizations

Let's explore the house price data. Last year, I made several *[Visual Meditations on House Prices]*({% post_url 2016-05-08-visual-meditations-on-house-prices %}) exploring different ways to look at house price trends.  Let's revisit one of my favorite meditations and render it with plotly.

## State dot plot

Let's explore trends in house prices, with a dot plot:

```{r jan-29-2017-hpi-graph-3,fig.height=10}

#compute rolling min/max of house price index
df.state<-df.state[, hpi12min:=rollapply(hpi, 12, min,fill=NA, na.rm=FALSE,align='right'), by=state]
df.state<-df.state[, hpi12max:=rollapply(hpi, 12, max,fill=NA, na.rm=FALSE,align='right'), by=state]

g3<-
  ggplot(data=df.state[(year>=2016) & month==9 & state !="DC" & state !="US"], aes(x=hpi, y=reorder(state,hpi), label=state,color=hpa12))+
  
  geom_text(nudge_x = 0.025)  +
  
  geom_point()+scale_x_log10(limits=c(70,350), breaks=c(70,100,150,250,350))+
  
  geom_segment(aes(xend=hpi12min,x=hpi12max,y=reorder(state,hpi),yend=reorder(state,hpi)),alpha=0.7)+
  theme_minimal()  +
  scale_colour_gradient(low="red",high="blue",name = "12-month HPA",labels = percent)+
  labs(y="", x="House price index (log scale, Dec 2000 =100, NSA)",
       title="State house price dynamics",
       subtitle=paste(as.character(df.state[year==2016 & month==9 & state=="US"]$date,format="%b-%Y")),
       caption="@lenkiefer Source: Freddie Mac house price index, each dot a state, lines trailing 12-month min-max")+
  theme(plot.title=element_text(size=18))+
  theme(plot.caption=element_text(hjust=0,vjust=1,margin=margin(t=10)))+
  theme(plot.margin=unit(c(0.25,0.25,0.25,0.25),"cm"),
        axis.text.y=element_blank())+ 
  
  theme(legend.justification=c(0,0), legend.position=c(.8,.75))
g3
```

Once again, with slith modifications we can render this chart (and animate it over years) with plotly.

```{r jan-29-2017-hpi-graph-3-plotly}
g3<-
  ggplot(data=df.state[(year>=2000) & month==9 & state !="DC" & state !="US"], aes(x=hpi, y=reorder(state,hpi), label=paste("  ",state,"  "), color=hpa12,frame=year,ids=state))+
  geom_text(nudge_x= 0.005,size=2)  +
  geom_point(alpha=0.82)+scale_x_log10(limits=c(70,350), breaks=c(70,100,150,250,350))+
  theme_minimal()  +
  scale_colour_gradient(low="red",high="blue",name = "12-month HPA",labels = percent)+
  labs(y="State", x="House price index (log scale, Dec 2000 =100, NSA)<br>Source: Freddie Mac House Price Index in September of year",
       title="State house price dynamics by @lenkiefer")+
  theme(plot.title=element_text(size=14),
        axis.text.y=element_blank())+
  theme(plot.caption=element_text(hjust=0,vjust=1,margin=margin(t=10)))+ 
  theme(legend.justification=c(0,0), legend.position=c(.8,.75))
ggplotly(g3) %>% animation_opts(frame=1500,transition=750,redraw=T)  
```

Once again, the graph above is a static screenshot, but you can see the interactive version by [clicking here]({{ site.url}}/chartbooks/jan2017/plotly/hpa-jan29-2017-1.html).

## Metro distribution

Let's explore one more plot. This one will show the distribution of annual house price percentage changes over the more than 300 metro areas tracked in the [Freddie Mac House Price Index](http://www.freddiemac.com/finance/house_price_index.html).

We want to see how the distribution of house price percent changes across metro areas has changed over time.  For September of each year in our sample we will find a kernel density function over metro house price percent changes. 
Then we'll examine the data using a static *ggplot2* graph.  Finally, we'll animate the plot using plotly.

### Fit a static plot

The code below takes our house price data and fits a kernel density function. We also create a couple of function `myf()` and `myf2()` to subset the data and stack a bunch of data frames using `rbind()`.

```{r jan-29-2017-hpi-graph-4}
df$year<-year(df$date)
df$month<-month(df$date)
myf<-function(yy){
  df.area<-data.frame(x=density(df[year==yy & month==9]$hpa12)$x,
                      y=density(df[year==yy & month==9]$hpa12)$y)
  df.area$year<-yy
  return(df.area)
}

# Create a function to stack the fitted densities in a data frame:

myf2<-function(start,end){
  my.out<-myf(start)
  for (i in (start+1):end){
    my.out<-rbind(my.out,myf(i))
  }
  return(my.out)
}

# Create plot:

g4<-
  ggplot(data=myf2(2005,2016),aes(x=x,y=y,fill=factor(year),group=year,frame=year))+geom_area(alpha=0.7,position="identity")+
  facet_wrap(~year)+
  scale_color_viridis(discrete=T)+ theme_minimal()+
  scale_fill_viridis(discrete=T)+scale_x_continuous(label=percent)+
  labs(x="Annual % Change in House Prices",y="density",
       title="Distribution of annual house price percent change",
       subtitle="over metro areas",
       caption="@lenkiefer Source: Freddie Mac House Price Index")+
  theme(plot.title=element_text(size=18),
        legend.position="none",
        plot.caption=element_text(hjust=0,vjust=1,margin=margin(t=10)),
        plot.background=element_rect(fill="#fffff8",color=NA))
  g4
```

We can see from this graph that house price appreciation was was more dispersed from 2005-2010 than in recent years.  Initially from 2005-2006 many metros had very high positive house price appreciation. Then, in the recession house prices fell quite a lot in mnay metros, but not all.  In recent years house prices have tended to be positive, but we don't see as many extremes as in 2005-2006.

The small multiple from using `facet_wrap()` with ggplot2 gives us one way of seeing this pattern.  But an animation might also help us to better understand.

### Animated density with plotly

We can modify this code slighlty to make a plotly animated chart. The following code using ggplotly to animate the house price densities we estimated above:

```{r jan-29-2017-hpi-graph-4-plotly}
# Create a function to fit a kernel density to house prices over the metro areas:
g4<-
  ggplot(data=myf2(2005,2016),aes(x=x,y=y,fill=factor(year),group=year,frame=year))+geom_area(alpha=0.7,position="identity")+
  scale_color_viridis(discrete=T)+ theme_minimal()+
  scale_fill_viridis(discrete=T)+scale_x_continuous(label=percent)+
  geom_text(x=-.25,y=12,hjust=0,aes(label=paste0("Sep. ",year),color=factor(year)),
            size=12,fontface="bold")+ theme(legend.position="none")+
  labs(x="Annual % Change in House Prices",y="density",
       title="Distribution of annual house price percent change<br>over metro areas<br>@lenkiefer Source: Freddie Mac House Price Index")
ggplotly(g4) %>% animation_opts(frame=1000,transition=600,redraw=T) 

```

Once again, the graph above is a static screenshot, but you can see the interactive version by [clicking here]({{ site.url}}/chartbooks/jan2017/plotly/hpa-jan29-2017-2.html).

# What's Next?

I have been having a lot of fun with plotly recently.  I've been using it with the flexdashboards I've been exploring, but it's also nice to use on its own. Lots more is left to explore.


